Efficient Compilation of High Level Python Numerical Programs with Pythran

### 2014-09-05 doc/papers/cgo2014/paper.tex
Fix dangling reference (in LaTeX :-)
CGO: partial Review

    This papers focuses on the case of the Python language and especially the
    Numpy package that provides core array data structure and basic linear
    algebra routines. It first conducts a case study based on Numpy use
    cases from the StackOverflow question and answer website, then studies
    existing compilers for numerical Python. Optimization opportunities,
    including parallelization and vectorization, and their implementation in
    the Pythran compiler are presented and illustrated through several
    benchmarks, showing very interesting speedups over the standard Python
    interpreter. Speedup greater than a factor of 10 are achieved despite the
    fact that the considered benchmarks mostly call Numpy routines that run C code.

@@

This paper focuses on compilation techniques that are relevant for the
optimization of high-level numerical kernels written in Python using the Numpy
package. Its shows an implementation of these techniques through the Pythran
compiler and provided a benchmarks test suite for performance measures with
others Python compilers.
Section~\ref{sec:stackoverflow} presents a case study based on user
input gathered on the StackOverflow question and answer site.
Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
showcases the optimization opportunities typically found in Numpy-based
kernels. Existing compilation approach to optimize such kernels are discussed
in Section~\ref{sec:compilers}, then Section~\ref{sec:pythran} emphasises on a
compiler-runtime cooperative approach to solve the performance issues.
Experimentations showcasing important performance gains are detailed in
Section~\ref{sec:xp}.

@@

StackOverflow~\footnote{\url{http://stackoverflow.com/}} is a question and
answer site for programmers. Any registered user can submit questions and other
users may answer. A voting system is then used to sort the answers. Said
otherwise, the website provides a database of question and answers on
programming topics, including Python.

A query using the terms \emph{numpy} and \emph{slow} yields a lot of results, many
of which use the pattern ``Q: my code is slow'' followed by an upvoted answer
``A: rewrite it that way'', where the original code performs explicit looping
over an array, and the rewritten code uses the relevant combination of Numpy
function.

@@

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in Numpy.}, label={lst:l2norm-loopy}]
def slow(x):
    r = np.empty(x.shape[0])
    for i in xrange(x.shape[0]):
        r[i] = np.sum(np.abs(x[i])**2)
    return r
\end{lstlisting}

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in Numpy.}, label={lst:l2norm}]
import numpy as np
def l2norm(x):
    return np.sum(np.abs(x)**2, axis=1)
\end{lstlisting}

Newcomers to Python with a prior C or Fortran experience are likely to make
use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their
command of the Numpy API and of good programming practices grows, they get to write code similar
to Listing~\ref{lst:l2norm}. The fact that in Numpy, high-level constructs run faster is also a
strong motivation. As a direct consequence, a compiler should focus on optimizing
high-level constructs rather than raw
loops. Both to encourage good practice with respect to Numpy code, and because
this is the kind of code Numpy users ultimately write.

@@

On the opposite, if all computations were merged into a single loop using
temporaries elimination and operator fusion, parallelization would be more
effective as barrier between binary operations are no longer needed and loads
and stores will be used to perform multiple operations.

@@

The limitation of expression templates presented in
\S~\ref{sec:expression-templates} can be solved using forward substitution at
the Python compiler level. This optimization leverages the
\textit{def-use} chains~\cite{defuse} as well as interprocedural read-write
effects and alias analysis presented in the next sections.
Thanks to these analyses, the forward substitution implementation is quite
simple: \todo{it is not THAT simple indeed, there is "lazyness" analysis
involved}....

@@

\subsection{Comparison with others Python compiler}
\todo{a better title have to be found}

@@

    \centering{
\begin{tabular}{|l||c|c|c|c|c|}
\hline
& Python & numba & parakeet & pypy & pythran \\
\hline
\hline
         allpairs\_distances &   58293 &             0 &          3170 &   59232 & \textbf{1583} \\
  allpairs\_distances\_loops &   64137 &         72748 &          3158 &   65287 & \textbf{3146} \\
               arc\_distance &    2796 &          2792 &          3084 &    2791 &  \textbf{103} \\
                        conv & 3260079 &             0 & \textbf{3284} & 3268298 &          3420 \\
                create\_grid &    5231 & \textbf{5044} &             0 &    5240 &          5587 \\
                    cronbach &    1905 &          2016 &             0 &    1909 & \textbf{1890} \\
                   diffusion &   42998 &         45030 & \textbf{8843} &   43238 &         21704 \\
                      evolve &    8236 &             0 &             0 &    8319 & \textbf{6242} \\
                        fdtd & 1940884 &       1498849 &             0 & 2007839 & \textbf{2294} \\
                         fft &   27120 &             0 &             0 &   27009 & \textbf{2151} \\
                    grouping &    3739 &             0 &             0 &    3740 & \textbf{1747} \\
                     growcut & 2112332 &         11617 & \textbf{5485} & 2141571 &         10338 \\
                      harris &    7173 &          9116 & \textbf{4240} &    7322 &          4889 \\
                     hyantes &  288852 &             0 & \textbf{2834} &  289074 &          3522 \\
                       julia & 1744357 &          5168 &  \textbf{109} & 1771329 &          4255 \\
                      l2norm &    5966 &          5948 &          9210 &    6559 &  \textbf{551} \\
               local\_maxima &  132237 &             0 &             0 &  132047 & \textbf{8561} \\
                      lstsqr &    9254 &          7528 &             0 &    8992 & \textbf{1414} \\
                      mandel &  951612 & \textbf{8175} &             0 &  951687 &         10341 \\
               multiple\_sum &    5477 &          5813 &          9805 &    5475 & \textbf{3503} \\
                    pairwise & 3362150 &         10764 &          6199 & 3347574 & \textbf{6015} \\
              periodic\_dist &    2559 &          2675 &             0 &    2562 & \textbf{2292} \\
                   repeating &    2808 &          2836 &             0 &    2795 &  \textbf{810} \\
             reverse\_cumsum &    4583 &          4828 &             0 &    4583 & \textbf{2379} \\
                       rosen &   15382 &         14309 &          2763 &   16065 &  \textbf{939} \\
                   slowparts &    8302 &             0 & \textbf{1610} &    8241 &          1927 \\
                   smoothing &  834057 &          8478 &          9771 &  819827 & \textbf{8254} \\
             specialconvolve &   12537 &         11978 &          6039 &   12570 & \textbf{5572} \\
                vibr\_energy &    5277 &          5275 &          5334 &    5416 &  \textbf{146} \\
                        wave &   72394 &         67590 &             0 &   72300 & \textbf{5186} \\
                       wdist &  108572 &        119211 &          4197 &  108283 & \textbf{4010} \\
\hline
\end{tabular}
}
\label{tbl:overview}
\caption{Average execution times for all kernels of the numpy-benchamrks on Intel8 with OpenMP (8 threads) and SIMD extension.}
\end{table*}

>>>>

    This papers focuses on the case of the Python language and especially the
    Numpy package that provides core array data structure and basic linear
    algebra routines. It first conducts a case study based on Numpy use
    cases from the StackOverflow questions and answers website, then studies
    existing compilers for numerical Python. Optimization opportunities,
    including parallelization and vectorization, and their implementation in
    the Pythran compiler are presented and illustrated through several
    benchmarks, showing very interesting speedups over the standard Python
    interpreter. Speedup greater than a factor of 10 are achieved despite the
    fact that the considered benchmarks mostly call Numpy routines that run C code.

@@

This paper focuses on compilation techniques that are relevant for the
optimization of high-level numerical kernels written in Python using the Numpy
package. Its introduces the implementation of these techniques in the Pythran
compiler and provides a benchmarks test suite for performance measures along
several Python compilers.
Section~\ref{sec:stackoverflow} presents a case study based on user
input gathered on the StackOverflow questions and answers website.
Section~\ref{sec:optimize} focuses on a simple yet representative kernel and
showcases the optimization opportunities typically found in Numpy-based
kernels. Existing compilation approaches to optimize such kernels are discussed
in Section~\ref{sec:compilers}. Section~\ref{sec:pythran} emphasises on a
compiler-runtime cooperative approach to solve the performance issues.
Finally, experimentations showcasing important performance gains are detailed in
Section~\ref{sec:xp}.

@@

StackOverflow~\footnote{\url{http://stackoverflow.com/}} is a question and
answer site for programmers. Any registered user can submit questions and other
users may answer. A voting system is then used to sort the answers. Said
otherwise, the website provides a database of questions and answers on
programming topics, including Python.

A query using the terms \emph{numpy} and \emph{slow} yields a lot of results, many
of which expose the pattern ``Q: my code is slow'' followed by an upvoted answer
``A: rewrite it that way'', where the original code performs explicit looping
over an array, and the rewritten code uses the relevant combination of Numpy
function.

@@

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm with loop in Numpy.}, label={lst:l2norm-loopy}]
def l2norm_slow(x):
    r = np.empty(x.shape[0])
    for i in xrange(x.shape[0]):
        r[i] = np.sum(np.abs(x[i])**2)
    return r
\end{lstlisting}

\begin{lstlisting}[language=python,caption={Per row version of $L^2$ norm without loop in Numpy.}, label={lst:l2norm}]
import numpy as np
def l2norm_fast(x):
    return np.sum(np.abs(x)**2, axis=1)
\end{lstlisting}

Newcomers to Python with a prior C or Fortran experience are likely to make
use of raw loops similar to Listing~\ref{lst:l2norm-loopy}, but as their
knowledge of the Numpy API and of good programming practices grows, they get to 
write code similar to Listing~\ref{lst:l2norm}. 
The fact that in Numpy, high-level constructs run faster is also a strong 
motivation. As a direct consequence, a compiler should focus on optimizing
high-level constructs rather than raw
loops. Both to encourage good practice with respect to Numpy code, and because
this is the kind of code Numpy users ultimately write.

@@

On the opposite, if all computations were merged into a single loop using
temporaries elimination and operator fusion, parallelization would be more
effective as barriers between binary operations are no longer needed and loads
and stores will be used to perform multiple operations.

@@

The limitation of expression templates presented in
\S~\ref{sec:expression-templates} can be solved using forward substitution at
the Python compiler level. This optimization leverages the
\textit{def-use} chains as well as interprocedural read-write
effects and alias analysis presented in the next sections.
Thanks to these analyses, the forward substitution implementation is quite
simple: \todo{it is not THAT simple indeed, there is "lazyness" analysis
involved}....

@@

\subsection{Comparison with others Python compiler}

@@

    \centering{
\begin{tabular}{|l||c|c|c|c|c|}
\hline
& Python & numba & parakeet & pypy & pythran \\
\hline
\hline
         allpairs\_distances &   58293 &             0 &          3170 &   59232 & \textbf{1583} \\
  allpairs\_distances\_loops &   64137 &         72748 &          3158 &   65287 & \textbf{3146} \\
               arc\_distance &    2796 &          2792 &          3084 &    2791 &  \textbf{103} \\
                        conv & 3260079 &             0 & \textbf{3284} & 3268298 &          3420 \\
                create\_grid &    5231 & \textbf{5044} &             0 &    5240 &          5587 \\
                    cronbach &    1905 &          2016 &             0 &    1909 & \textbf{1890} \\
                   diffusion &   42998 &         45030 & \textbf{8843} &   43238 &         21704 \\
                      evolve &    8236 &             0 &             0 &    8319 & \textbf{6242} \\
                        fdtd & 1940884 &       1498849 &             0 & 2007839 & \textbf{2294} \\
                         fft &   27120 &             0 &             0 &   27009 & \textbf{2151} \\
                    grouping &    3739 &             0 &             0 &    3740 & \textbf{1747} \\
                     growcut & 2112332 &         11617 & \textbf{5485} & 2141571 &         10338 \\
                      harris &    7173 &          9116 & \textbf{4240} &    7322 &          4889 \\
                     hyantes &  288852 &             0 & \textbf{2834} &  289074 &          3522 \\
                       julia & 1744357 &          5168 &  \textbf{109} & 1771329 &          4255 \\
                      l2norm &    5966 &          5948 &          9210 &    6559 &  \textbf{551} \\
               local\_maxima &  132237 &             0 &             0 &  132047 & \textbf{8561} \\
                      lstsqr &    9254 &          7528 &             0 &    8992 & \textbf{1414} \\
                      mandel &  951612 & \textbf{8175} &             0 &  951687 &         10341 \\
               multiple\_sum &    5477 &          5813 &          9805 &    5475 & \textbf{3503} \\
                    pairwise & 3362150 &         10764 &          6199 & 3347574 & \textbf{6015} \\
              periodic\_dist &    2559 &          2675 &             0 &    2562 & \textbf{2292} \\
                   repeating &    2808 &          2836 &             0 &    2795 &  \textbf{810} \\
             reverse\_cumsum &    4583 &          4828 &             0 &    4583 & \textbf{2379} \\
                       rosen &   15382 &         14309 &          2763 &   16065 &  \textbf{939} \\
                   slowparts &    8302 &             0 & \textbf{1610} &    8241 &          1927 \\
                   smoothing &  834057 &          8478 &          9771 &  819827 & \textbf{8254} \\
             specialconvolve &   12537 &         11978 &          6039 &   12570 & \textbf{5572} \\
                vibr\_energy &    5277 &          5275 &          5334 &    5416 &  \textbf{146} \\
                        wave &   72394 &         67590 &             0 &   72300 & \textbf{5186} \\
                       wdist &  108572 &        119211 &          4197 &  108283 & \textbf{4010} \\
\hline
\end{tabular}
}
\caption{Average execution times for all kernels of the numpy-benchamrks on Intel8 with OpenMP (8 threads) and SIMD extension.}
\label{tbl:overview}
\end{table*}
